---
title: "DATA 607 Week 1"
author: "Vinicio, Haro"
date: "February 2, 2018"
output:
  html_document: default
  pdf_document: default
---

1) Reading in the data and naming the columns 
read in the data from the URL
https://archive.ics.uci.edu/ml/machine-learning-databases/mushroom/agaricus-lepiota.data

Description of the data here: https://archive.ics.uci.edu/ml/datasets/Mushroom

The data does not contain any headers so set header=FALSE
Details here: https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html
```{r setup, include=FALSE}

library(data.table)
df.mushroom<-fread("https://archive.ics.uci.edu/ml/machine-learning-databases/mushroom/agaricus-lepiota.data", header= FALSE, sep=",")


```

Check the data 

```{r mushroom}
head(df.mushroom)
```

Now we need to use the provided documentation in order to name the columns with the correct names.
Lets create a vector that has all the column names. We can append this vector to the columns to assign
column names
This handy tutorial details the process I used:
http://rprogramming.net/rename-columns-in-r/
```{r pressure, echo=FALSE}
mushroom.col <- c("class", "capshape", "cap_surface","cap-color", "bruises?", "odor", "gill-attachment",
                    "gill-spacing", "gill-size", "gill-color", "stalk-shape" ,"stalk-root", "stalk-surface-above-ring",
                    "stalk-surface-below-ring", "stalk-color-above-ring", "stalk-color-below-ring" ,"veil-type",
                    "veil-color", "ring-number", "ring-type", "spore_print_color", "population", "habitat")

```

Append the column names to the dataframe 

```{r column names}
colnames(df.mushroom) <- mushroom.col
```

2) Select the column called class and an addition 3 or 4 columns. We need to create a subset of the data 
Why did I pick these columns? 
Lets say I wanted to find a relationship between class and some physical attributes. 
I could answer the question, can we predict if a mushroom is edible or poisonous by some of its 
physcal attributes?
Here we just use very basic commans to make the subset. We are not using any additional critera for more filtering. 
```{r}
df.mushroomb<-subset(df.mushroom, select=c('class', 'capshape', 'odor', 'spore_print_color'))
```

3) We need transform the columns we picked from abbreviated names to full names based on the data documentation.
Transform the class variable: map P to poisonous and e to edible as given in the documentation.

We can do this manually but a loop to iterate on each entry might be easier.
Lets convert the other entries in the selected columns to their full names as well. 
Refer to this link for a handy tutorial I used to write the following loops https://www.datamentor.io/r-programming/if-else-statement


```{r class mapping}
i <- 1
for (x in df.mushroomb$class)
{
  if (x == 'p')
  	  {
      df.mushroomb$class[i] <- "poisonous"
      } 
  	  else if (x == 'e')
  	  {
      df.mushroomb$class[i] <- "edible"
      }
  i <- i + 1
}


```

```{r capshape mapping }
i <- 1
for (x in df.mushroomb$capshape)
{
  if (x == 'b')
  	   {
    df.mushroomb$capshape[i] <- "bell"
       }
  	else if (x == 'c')
  	   {
    df.mushroomb$capshape[i] <- "conical"
       }
  	else if (x == 'x')
  	   {
    df.mushroomb$capshape[i] <- "convex"
       }
  	else if (x == 'f')
  	   {
    df.mushroomb$capshape[i] <- "flat"
       }
  	else if (x == 'k')
  	   {
    df.mushroomb$capshape[i] <- "knobbed"
       }
  	else if (x == 's'){
    df.mushroomb$capshape[i] <- "sunken"
       }
  i <- i + 1
}
```



```{r odor mapping}
i <- 1
for (x in df.mushroomb$odor)
{
  if (x == 'a')
  	   {
    df.mushroomb$odor[i] <- "almond"
       }
  	else if (x == 'l')
  	   {
    df.mushroomb$odor[i] <- "anise"
       }
  	else if (x == 'c')
  	   {
    df.mushroomb$odor[i] <- "creosote"
       }
  	else if (x == 'y')
  	   {
    df.mushroomb$odor[i] <- "fishy"
       }
  	else if (x == 'f')
  	   {
    df.mushroomb$odor[i] <- "foul"
       }
  	else if (x == 'm')
  	   {
    df.mushroomb$odor[i] <- "musty"
       }
       else if (x == 'n')
  	   {
    df.mushroomb$odor[i] <- "none"
       }
       else if (x == 'p')
  	   {
    df.mushroomb$odor[i] <- "pungent"
       }
       else if (x == 's')
  	   {
    df.mushroomb$odor[i] <- "spicy"
       }
  i <- i + 1
}
```

```{r spore print color mapping }
i <- 1
for (x in df.mushroomb$spore_print_color)
{
  if (x == 'k')
  	   {
    df.mushroomb$spore_print_color[i] <- "black"
       }
  	else if (x == 'n')
  	   {
    df.mushroomb$spore_print_color[i] <- "brown"
       }
  	else if (x == 'b')
  	   {
    df.mushroomb$spore_print_color[i] <- "buff"
       }
  	else if (x == 'h')
  	   {
    df.mushroomb$spore_print_color[i] <- "chocolate"
       }
  	else if (x == 'r')
  	   {
    df.mushroomb$spore_print_color[i] <- "green"
       }
  	else if (x == 'o')
  	   {
    df.mushroomb$spore_print_color[i] <- "orange"
       }
       else if (x == 'u')
  	   {
    df.mushroomb$spore_print_color[i] <- "purple"
       }
       else if (x == 'w')
  	   {
    df.mushroomb$spore_print_color[i] <- "white"
       }
       else if (x == 'y')
  	   {
    df.mushroomb$spore_print_color[i] <- "yellow"
       }
  i <- i + 1
} 
```

Troubleshooting: You may need to specify factors as false when you load in the data, 
otherwise the loops will not work. This depends on how the data is loaded in. fread seemed to fix it. 
How is our data so far? 
```{r new data frame}
head(df.mushroomb)
```
Data looks good! 
Looks good, lets explore the data 
What are the relative frequencies for the class variable?
Lets try some things to answer this question 

Get the counts of edible or poisonous 
```{r counts of edible/poisonous mushrooms}
table(df.mushroomb$class)
```
convert counts to proportions instead 
```{r proportions}
class_counts <- table(df.mushroomb$class)
class_counts / sum(class_counts)
```


lets visually summerize the data using a standard barplot 
We will use ggplot2 and then use summary to see binning and plotting details 
```{r plot}
library(ggplot2)
bar_plt <- ggplot(df.mushroomb, aes(x = class)) 
bar_plt <- bar_plt + geom_bar()
summary(bar_plt)
```


```{r, echo=FALSE}
bar_plt
```


There are many more variations of data exploration but for the task at hand, the data has been subsetted and renamed for downstream analysis. 

One last check:
```{r}
head(df.mushroomb)
```

```{r}
tail(df.mushroomb)
```

